Reasonably Programmable Syntax

### 2017-06-25 popl18/syntax-popl18.tex
revisions to introduction and abstract

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
%Many languages define convenient list literal forms. This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language? 
When designing the surface syntax of a programming language, it is common practice to build in literal forms for a few standard data structures, e.g. lists. This practice neglects an unbounded variety of other data structures for which literal forms might also be useful (consider, for example,  maps, regexes, HTML data, SQL queries and chemical structures.) %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.)% Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, defeating much of the purpose of convenient literal forms. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
This paper introduces a system of \emph{typed literal macros (TLMs)} that gives library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} at any specified type or parameterized family of types. % The mechanism is {strictly hygienic}, meaning that 1) the expansion must be context-independent; and 2)  splicing is capture avoiding. 
% Partial parameter application lowers the syntactic cost of this strict style. 
Compared to existing approaches, including syntax extension systems and macro-based quotation systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic determinism, and TLM clients can reason abstractly about the type and binding structure of the program. The system needs only to convey to clients (e.g. via secondary notation) the inferred \emph{segmentation} of each literal, which gives the locations of the spliced terms. The inferred segmentation also informs the hygiene mechanism (previously, only unhygienic macros could freely parse terms out of literal bodies.) 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
% This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules. This is the  first detailed type-theoretic account of a typed, hygienic term-rewriting macro system for an ML-like language. % This calculus is the first such formalization of a typed macro system.

@@

% \vspace{-4px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
The surface syntax of a programming language is a user interface. As such, language designers often include in the surface syntax various \emph{literal forms}  that are designed to decrease the cognitive cost of constructing and pattern matching over values of certain otherwise ordinary types from the language's standard library. For example, in Standard ML (SML) and its cousins, the list expression literal \li{[x, y, z]} desugars to a term equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the expansion of literal forms is \emph{context independent}. We will return to the concept of context independence again.} List pattern literals are analagous. Another example is Ur/Web, which extends Ur's textual syntax with expression and pattern literals for encodings of XHTML data, as defined in Ur/Web's standard library \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Fig. \ref{fig:urweb} shows examples of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2). 
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

Third-party library providers are quite justifiably envious of the special syntactic status of these semantically ordinary standard library constructs. After all, it is not difficult to find other examples of types for which mathematicians, scientists and programmers have invented specialized notations. For example, 1) clients of a ``collections'' library might like not just list literals, but also matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just XHTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}. \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use string encodings and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}. Choosing strings over composite data structures for the sake of convenience, a.k.a. ``stringly-typed programming'', is rather common in practice -- \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar, e.g. for SQL queries, regexes, file system paths, URLs and others. Literal forms at these types, with support for splicing, would lower the cost of richer compositional encodings.%Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

\begin{figure}
\vspace{-2px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-10px}
\caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web.}
\vspace{-7px}
\label{fig:urweb}
\end{figure}

Language designers cannot, of course, hope to accomodate all such types with primitive literal forms \emph{a priori}, so there has been considerable interest in mechanisms that allow programmers to define new literal forms on their own. For example, we will describe a number of syntax extension mechanisms, including Camlp4 \cite{ocaml-manual} and SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, in Sec. \ref{sec:existing-approaches}. The problem is that these mechanisms quickly become \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
Consider the perspective of a programmer attempting to comprehend (i.e. reason about) the program text in Fig. \ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax where a large number of syntax extensions defining novel literal forms, some unfamiliar to the programmer, are active.

\begin{figure}[h]
\vspace{-10px}
\begin{lstlisting}
  val w = compute_w()
  val x = compute_x(w)
  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
\end{lstlisting}
\vspace{-10px}
\caption{An example of unreasonable program text.}
\vspace{-9px}
\label{fig:K-dialect}
\end{figure}

\noindent
The literal form on Line 3 constructs an encoding of a query in the stack-based database query language K, using its terse textual syntax \cite{Whitney:2001:LOR:376284.375783}. Of course, the programmer reading this program text may well be unfamiliar with K's query syntax. The problem then is that there are no simple, abstract protocols for answering questions like:

\begin{enumerate}
\item \textbf{(Responsibility)} Which syntax extension determined the desugaring of the literal form on Line 3? Might there be many possible desugarings (i.e. syntactic non-determinism)?
\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 taken as spliced expressions (meaning that they appear directly in the desugaring), or are they parsed in some other way peculiar to this literal form (e.g. as operators in the K language)?
\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier used in the desugaring of Line 3?
\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break Line 3, or change its meaning because the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
\item \textbf{(Typing)} What type does \li{y} have?
\end{enumerate}

% In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. 
Forcing the programmer to reason transparently, i.e. to examine the desugaring of the program text, or the logic that produced that desugaring, in order to answer these sorts of basic questions,  can defeat the purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}. Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding, and types mediate abstraction over function and module implementations.  %From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.

This paper details a more reasonable mechanism: \emph{typed literal macros (TLMs)}. TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e. they are initially parsed like raw string literals.) For example, Fig. \ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a backtick-delimited generalized literal form to express the ``unreasonable'' K query example from Fig. \ref{fig:K-dialect}. The applied TLM parses the literal body and generates an expansion during a semantic phase called typed expansion, which generalizes the usual typing phase.  TLMs come equipped with uniquely powerful abstract reasoning principles. We will characterize these reasoning principles precisely as we proceed, but for now, let us develop some intuitions -- the programmer can reason abstractly about Line 3 as follows:
\begin{enumerate}
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for parsing and expansion of the literal body. The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax this TLM implements is deterministic.
\item \textbf{(Segmentation)} The output that a TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM. The segmentation can be communicated by a code editor using secondary notation, e.g. colors in this document. By examining Fig. \ref{fig:K-tsm-example}, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way (because they are in green.)
\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expressions \li{x} must refer to the binding of \li{x} on Line 2. They cannot capture a coincidental binding of \li{x} in the expansion.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}.
\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have. We will see an example of a TLM definition in the next section. Moreover, each spliced segment in the segmentation also comes paired with the type it is expected to have. This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

@@

\begin{figure}
\vspace{-4px}
\begin{lstlisting}
  val w = compute_w()
  val x = compute_x(w)
  val y = $kquery `SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL`
\end{lstlisting}
\vspace{-10px}
\caption{TLMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
\vspace{-12px}
\label{fig:K-tsm-example}
\end{figure}

The remainder of this document is organized as follows.
% \begin{itemize}
% \item 
Sec. \ref{sec:setsms} defines a {simple expression TLM} for XHTML, which implements Ur/Web's XHTML literal syntax, and describes the typed expansion process in more detail.
% \item 
Sec. \ref{sec:sptsms} briefly introduces {simple pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec. \ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above.
% \item 
Sec. \ref{sec:ptsms} adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that take type and module parameters. Parameters serve two purposes:
\begin{enumerate}[noitemsep,nolistsep]
\item They allow a TLM to operate not just at a single specified type, but over a type- and module-parameterized family of types. We describe a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type.
\item Parameters also give expansions access to terms other than those explicitly spliced in to the literal body by the client. %This is particularly convenient for working with abstract types \emph{a la} ML. 
Partial parameter application is possible, which lowers the syntactic cost of the strict context independence discipline that TLMs enforce.
\end{enumerate}
This section culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with type functions, ML-style modules and parametric TLMs.
% \item
Sec. \ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e. \emph{static evaluation}, in  more detail, and also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.
% \item 
Sec. \ref{sec:existing-approaches} compares TLMs to related work, including closely related work by \citet{TSLs,sac15} on type-specific languages, and the long history of work on macro systems.
% \item 
Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of the present limitations and some future research directions.
% \end{itemize}

Due to space limitations, many standard or straightforward technical details of the calculi defined in Sec. \ref{sec:setsms-formally} through Sec. \ref{sec:static-eval} are relegated to the supplemental material. The examples that we give are written in an emerging dialect of ML called VerseML that supports TLMs as described herein, but the essence of the mechanism is not ML-specific -- it should be straightforward to adapt it to other languages that take a similar approach to types and binding (e.g. Haskell, Scala and others.) When a design choice is ML-specific, we will make note of it\todo{do this}. Languages that have a disciplined binding structure but that lack a rich static type structure, e.g. Racket and other LISP derivatives, may also benefit from TLMs -- they can apply the results of this paper by deploying the usual type theorist's trick of viewing such a language as ``{unityped}'' \cite{pfpl,scott1980lambda}. % For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs. %VerseML is, as its name suggests, a conceptual descendent of ML. It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}. 
%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work. 
% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
% All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.

@@

We will begin in this section by introducing \emph{simple expression TLMs} (seTLMs), i.e. TLMs where expansions are expressions of a single specified type, by example.

Our take on the running example is shown in Fig. \ref{fig:first-tsm-example}: we apply a TLM named 
\li{$html} 
 to \emph{generalized literal forms} delimited by backticks. TLM names are prefixed by \li{$} 
 to clearly them from variables. The semantics delegates control over the parsing and expansion of each literal body to the applied TLM during a semantic phase called \emph{typed expansion}, which generalizes the usual typing phase. 
\begin{figure}[ht!]
\vspace{-10px}
\begin{lstlisting}
  fun heading first_name = $html `SURL<h1>Hello, {[EURLfirst_nameSURL]}!</h1>EURL`
  val body = $html `SURL<body>{EURLheading "World!"SURL} ...</body>EURL`
\end{lstlisting}
\vspace{-10px}
\caption{Two examples of a TLM being applied to a generalized literal form. %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase. 
We assume strings are primitive.}
\vspace{-10px}
\label{fig:first-tsm-example}
\end{figure}

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
%Many languages define convenient list literal forms. This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language? 
When designing the surface syntax of a programming language, it is common practice to build in literal forms for a few standard data structures, e.g. lists. This practice is \emph{ad hoc} -- there are an unbounded variety of other data structures for which literal forms might also be useful (consider, for example,  maps, regexes, HTML data, SQL queries, syntax trees and chemical structures.) %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.)% Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, defeating much of the purpose of convenient literal forms. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
This paper introduces a system of \emph{typed literal macros (TLMs)} that gives library providers full programmatic control over the parsing and expansion of expressions and patterns of \emph{generalized literal form} at \emph{any} specified type or parameterized family of types. % The mechanism is {strictly hygienic}, meaning that 1) the expansion must be context-independent; and 2)  splicing is capture avoiding. 
% Partial parameter application lowers the syntactic cost of this strict style. 
Compared to existing approaches, including syntax extension systems and macro-based quotation systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic determinism, and TLM clients can reason abstractly about the type and binding structure of the program. The system only needs to convey to clients (via secondary notation) the inferred \emph{segmentation} of each literal, which gives the locations of the spliced terms. The inferred segmentation also informs the hygiene mechanism (previously, only unhygienic macros could freely parse terms out of string literal bodies.) 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
% This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules. This is the  first detailed type-theoretic account of a typed, hygienic term-rewriting macro system of any design for an ML- or Scala-like language. % This calculus is the first such formalization of a typed macro system.

@@

% \vspace{-4px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is common practice to build in a few \emph{literal forms} that lower the syntactic cost of constructing and pattern matching over values of certain standard types. For example, in Standard ML (SML) and its cousins, the list expression literal \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. We will return to the concept of context independence again.} List pattern literals are analagous. Another example is found in Ur/Web's surface syntax, which extends Ur's surface syntax with expression and pattern literal forms for encodings of XHTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Fig. \ref{fig:urweb} gives examples of XHTML literals that ``splice in'' a string expression (Line 1) and another XHTML expression (Line 2). 
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

This design practice is decidedly \emph{ad hoc} -- it is easy to find other examples of  library constructs that are not privileged with literal forms by these languages, but for which mathematicians, scientists or programmers have invented and adopted specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}. Consider, for example, that 1) clients of a ``collections'' library might like not just list literals, but also matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just XHTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like composite chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.

Although concerns about literal notation are easy to dismiss as superficial, the reality is that literal forms, or the absence thereof, have had a substantial influence on the quality of software that is written today. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal forms for structured encodings of queries, like the SQL query literals found in many languages \cite{LINQ}\todo{find citation}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}. More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e. they choose strings instead of composite data structures largely for reasons of syntactic convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them were parseable according to some readily apparent type-specific grammar, e.g. for SQL queries, regexes, file system paths, URLs and many other structures. Literal forms at these types, with support for splicing, would decrease the syntactic cost of composite encodings, which are easier than flat string encodings to manipulate programmatically and to reason about compositionally. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

\begin{figure}
\vspace{-2px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-10px}
\caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web.}
\vspace{-8px}
\label{fig:urweb}
\end{figure}

Of course, it would not be sensible to ask general-purpose language designers to accomodate all known notations with primitive literal forms. Instead, there has been considerable interest in mechanisms that allow library providers to define new literal forms on their own. For example, general syntax extension mechanisms, including Camlp4 \cite{ocaml-manual}, SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and others that we will return to in Sec. \ref{sec:existing-approaches}, can be used to define new literal forms. The problem is that these mechanisms reveal themselves to be quite \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
Consider, for instance, the perspective of a programmer attempting to comprehend (i.e. reason about) the program text in Fig. \ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax where a large number of new literal forms, some unfamiliar to the programmer, have been defined using one of these mechanisms.

\begin{figure}[h]
\vspace{-10px}
\begin{lstlisting}
  val w = compute_w()
  val x = compute_x(w)
  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
\end{lstlisting}
\vspace{-10px}
\caption{An example of unreasonable program text.}
\vspace{-10px}
\label{fig:K-dialect}
\end{figure}

\noindent
The literal form on Line 3 above constructs an encoding of a query in the stack-based database query language K, using its terse surface syntax \cite{Whitney:2001:LOR:376284.375783}. Of course, the programmer reading this program text may well be unfamiliar with K's query syntax. Unfortunately for the programmer, there are no simple, abstract protocols for answering questions like:

\begin{enumerate}[nolistsep,noitemsep]
\item \textbf{(Responsibility)} Which syntax extension determined the desugaring of the literal form on Line 3? Might there be many possible desugarings (i.e. syntactic non-determinism)?
\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 taken as spliced expressions (meaning that they appear directly in the desugaring), or are they parsed in some other way peculiar to this literal form (e.g. as operators in the K language)?
\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier used in the desugaring of Line 3?
\item \textbf{(Context Dependence)} If \li{w}, on Lines 1-2, is renamed, could that possibly break Line 3 or change its meaning because the desugaring assumes that some variable identified as \li{w} is in scope? What other hidden assumptions might the desugaring be making?
\item \textbf{(Typing)} What type does \li{y} have?
\end{enumerate}

% In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. 
Forcing the programmer to reason transparently, i.e. to examine the desugaring of the program text, or even the logic that produced that desugaring, defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}. Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding, and types mediate abstraction over function and module implementations. % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.

This paper formally details a more reasonable mechanism for defining new literal forms: \emph{typed literal macros (TLMs)}. TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms} \cite{TSLs}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e. they are initially parsed like raw string literals.) For example, Fig. \ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a backtick-delimited generalized literal form to express the ``unreasonable'' K query example from Fig. \ref{fig:K-dialect}. The applied TLM parses the literal body and generates an expansion statically, during a semantic phase called typed expansion, which generalizes the usual typing phase.  TLMs come equipped with uniquely powerful abstract reasoning principles. We will characterize these reasoning principles absolutely precisely as we proceed, but for now, let us develop some intuitions: the programmer can reason abstractly about Line 3 as follows:
\begin{enumerate}[noitemsep,nolistsep]
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for parsing and expansion of the literal body. The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax this TLM implements is deterministic.
\item \textbf{(Segmentation)} The output that a TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM. The segmentation can be communicated by a code editor using secondary notation, e.g. colors in this document. By examining Fig. \ref{fig:K-tsm-example}, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way (because they are in green.)
\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expressions \li{x} must refer to the binding of \li{x} on Line 2. They cannot capture a coincidental binding of \li{x} in the expansion.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}.
\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have. (We will see an example of a TLM definition in the next section.) Moreover, each spliced segment in the segmentation also comes paired with the type it is expected to have. This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

@@

\begin{figure}
\vspace{-2px}
\begin{lstlisting}
  val w = compute_w()
  val x = compute_x(w)
  val y = $kquery `SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL`
\end{lstlisting}
\vspace{-10px}
\caption{TLMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
\vspace{-12px}
\label{fig:K-tsm-example}
\end{figure}

The remainder of this document is organized as follows.
% \begin{itemize}
% \item 
Sec. \ref{sec:setsms} defines a {simple expression TLM} that implements Ur/Web's XHTML literal syntax, and describes the typed expansion process in more detail.
% \item 
Sec. \ref{sec:sptsms} briefly introduces {simple pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec. \ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above.
% \item 
Sec. \ref{sec:ptsms} then adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that take type and module parameters. Parameters serve two purposes. First, they allow a TLM to operate not just at a single specified type, but over a type- and module-parameterized family of types. We describe a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type. Second, parameters give expansions access to terms other than those explicitly spliced in to the literal body by the client. %This is particularly convenient for working with abstract types \emph{a la} ML. 
Partial parameter application is possible, which lowers the syntactic cost of the strict context independence discipline that TLMs enforce. % Sec. \ref{sec:ptsms} culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with type functions, ML-style modules and parametric TLMs.
% \item
Sec. \ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e. \emph{static evaluation}, in  more detail, and also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.
% \item 
Sec. \ref{sec:existing-approaches} compares TLMs to related work, including recent related work by \citet{TSLs,sac15}, and the long history of work on macro systems.
% \item 
Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of present limitations and future work.
% \end{itemize}

Due to space limitations, many technical details of the calculi defined in Sec. \ref{sec:setsms-formally} through Sec. \ref{sec:static-eval}, as well as proofs, are relegated to the supplemental material. Examples are written in an emerging dialect of  ML called VerseML that is focused on making (Standard) ML more ergonomic. However, the essence of the mechanism is not ML-specific -- it should be straightforward to adapt it to other languages that take a similar approach to types and binding (e.g. Haskell, Scala and others.) When a design choice is ML-specific, we will make note of it\todo{do this}. Languages that have a disciplined binding structure but that lack a rich static type structure, e.g. Racket and other LISP derivatives, would also benefit from TLMs -- they can apply the results of this paper by deploying the usual type theorist's trick of viewing such a language as ``{unityped}'' \cite{pfpl,scott1980lambda}. % For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs. %VerseML is, as its name suggests, a conceptual descendent of ML. It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}. 
%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work. 
% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
% All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.

@@

We will begin in this section with another example that informally introduces \emph{simple expression TLMs} (seTLMs), i.e. TLMs for generating expressions of a single specified type.
